---
title: "Forecasting Foreign Aid Amounts"
output: pdf_document
---

This is a walk through report explaining the process of using forecast methodologies and apply those methodologies to financial foreign aid. This walk through is broken up into different sections:

* Introduction

* The Data Workflow
    + Data Import
    + Exploratory Data Analysis

* Forecast
    + Methodologies
    + Implementation

* Conclusions

#Introduction

Developing countries face challenges in accessing up-to-date information about aid, development, and humanitarian flows. This information is needed to plan and manage resources effectively. There is a large movement which allows different stakeholders to publish data on development cooperation activities. The International Aid Transparency Initiative [IATI](http://www.aidtransparency.net/) manages and provides the standard that donors and recipient countries, civil society organizations, and others use to publish and consume data on development cooperation activities.

Assumptions:

* All values are in USD

#Data Workflow

The first step of any analysis is to import the necessary packages. Packages used in this analysis are below.

```{r, results='hide', message=FALSE, warning=FALSE}
library(ggplot2) #static visualization
library(dplyr) #data manipulation and connect to database
library(knitr)
library("devtools") #needed to install plotly if you don't have it already

library(plotly) #interactive visualization
```

```
#needed for plotly API, replace it with yours, it's free if you sign up for an account
py <- plotly(username="XXXX", key="XXXX")
set_credentials_file("XXXXX", "XXXXX") 
```

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
#needed for plotly API, replace it with yours, it's free if you sign up for an account
py <- plotly(username="bcarancibia", key="1960rm8f1b")
set_credentials_file("bcarancibia", "1960rm8f1b") 
```
##Importing the CSV

The data was downloaded via the [IATI API](datastore.iatistandard.org/api/1/access/transaction.csv?recipient-country=BF%7CBI%7CTD%7CCD%7CCI%7CET%7CGM%7CGW%7CHT%7CHN%7CKG%7CLA%7CMG%7CMD%7CNP%7CSS%7CTL%7CUG&stream=True). The resulting csv was imported into memory to quickly view.

```{r}
transactions <- read.csv("transaction.csv", header=TRUE)
nrow(transactions)
```

The number of rows is quite large so the first step is to cut it down to only include transactions that are disbursments and commitments. In the transaction.type, fields with C, D, 2, or 3 are commitments and disbursements. First step is to subset the data to see how many rows of data are in the file.

```{r}
data <- subset(transactions, transaction.type =="C" | transaction.type =="D" | transaction.type ==2 | transaction.type == 3)
nrow(data)

nrow(transactions) - nrow(data)
```

The next step is to filter out only the necessary fields. Currently, there are 74, the key fields necessary are:

* transaction date
* default currency
* transaction value
* IATI identifier
* Reporting Org
* Title
* Start Planned
* End Planned
* Start Actual
* End Actual
* Recipient Country
* Sector
```{r}
data <- data[c("transaction.date","transaction.value", "iati.identifier", "reporting.org", "title", "start.planned", "end.planned", "start.actual", "end.actual", "recipient.country", "sector")]
```

The next step is to remove data that has multiple countries in the transactions, see example below.

```{r}
example.bad <- subset(data, recipient.country =="Pakistan;India;Nepal")
knitr::kable(head(example.bad))
```

I could split the transaction into three separate transactions and divide it equally amongst the three recipients, but this would be disengenious to the data and a data quality issue. To limit the scope of this study, the scope is limited to only transactions that are directly to a singular country.

```{r}
data$recipient.country<- as.character(data$recipient.country)
data$recipient.country[data$recipient.country == "Lao People'S Democratic Republic"] <- "Laos"


data <- subset(data, recipient.country=="Chad"|recipient.country=="Ethiopia"|recipient.country=="Nepal"|recipient.country=="Honduras"|recipient.country=="Senegal"|recipient.country=="Haiti"|recipient.country=="Burkina Faso"| recipient.country=="Burkina Faso"|recipient.country=="Tanzania, United Republic Of"|recipient.country=="Laos"|recipient.country=="Malawi"|recipient.country=="Moldova, Republic of"| recipient.country=="Madagascar"|recipient.country=="Uganda"|recipient.country=="Timor-Leste"|recipient.country=="Congo, The Democratic Republic Of The"|recipient.country=="Kyrgyzstan"|recipient.country=="")

data <- subset(data, transaction.value!="NA")
data <- subset(data, transaction.date!="")

```

Write data to csv and import it to sqlite.

```{r}
write.csv(data, file = "transactions.csv", row.names=TRUE, na="")
```







